---
title: "Data Preprocessing and Visualization"
subtitle: "<h2><u>Exploratory Data Analysis(UKBB40K)</u></h2>"
author: "<h3>Zhuoyi Yang (SOCR)</h3>"
date: "`r format(Sys.time(), '%B %Y')`"
tags: [Data Preprocessing, Dimensionality Reduction, Visualization, Feature Selection] 
output:
  html_document:
    theme: spacelab
    highlight: tango
    includes:
      before_body: SOCR_header.html
    toc: true
    number_sections: true
    toc_depth: 2
    toc_float:
      collapsed: false
      smooth_scroll: true
    code_folding: show
    self_contained: yes
---

# Packages

```{r packages, message=FALSE}
library(utility)
library(progress)
library(data.table)
library(Rtsne)
library(ggplot2)
library(plotly)
library(lubridate)
library(dummies)
library(umap)
library(dplyr)
library(missForest)
```

# Data Processing

total_cols: 22940L
total rows: 502493

```{r data statistics(columns and rows), message=FALSE}
initial_read <- fread("ukbb_data/ukb44534_compiled_tab-001.csv", nrows = 1)
total_cols <- ncol(initial_read)
total_rows <- as.numeric(system("wc -l < ukbb_data/ukb44534_compiled_tab-001.csv", intern=TRUE))
```

This code block sets up parallel processing capabilities, defines file paths, and initializes progress tracking. It includes a function to process each column of the dataset, determining the data type and saving summaries.

```{r data statistics(types)}
library(data.table)
library(foreach)
library(doParallel)

cores <- detectCores() - 1
registerDoParallel(cores)

data_path <- "ukbb_data/1000_rows.csv"
progress_path <- "progress_log.csv"
total_cols <- 29L

plot_dir <- paste0("plots/")
if (!dir.exists(plot_dir)) {
  dir.create(plot_dir, recursive = TRUE)
  }

summary_dir <- paste0("summaries/")
if (!dir.exists(summary_dir)) {
  dir.create(summary_dir, recursive = TRUE)
  }
if (file.exists(progress_path)) {
  progress <- fread(progress_path)
  start_col <- max(progress$col_processed) + 1
} else {
  fwrite(data.table(col_processed = integer(0)), progress_path)
  start_col <- 2  
}

process_column <- function(col) {
  column_titles <- fread(data_path, nrows = 1)
  col_title <- names(column_titles)[col]
  current_col_sample <- fread(data_path, select = col, nrows = 30)
  detected_type <- "Other"
  sample_values <- na.omit(current_col_sample[[1]])
  if (length(sample_values) == 0) {
    detected_type <- "Empty"
  } else if (all(is.numeric(sample_values))) {
    if (all(sample_values %in% c(0, 1))) {
      detected_type <- "Binary"
    } else {
      detected_type <- "Numeric"
    }
  } else if (any(grepl("^\\d{4}-\\d{2}-\\d{2}$", sample_values))) {
    detected_type <- "Date"
  } else {
    detected_type <- "Categorical"
    current_col <- fread(data_path, select = col)
    freq <- table(current_col[[1]], useNA = "ifany")
    freq <- as.data.frame(freq)
    freq$Prop <- freq$Freq / sum(freq$Freq)
    fwrite(freq, paste0("summaries/", col_title, ".csv"))
  }

  fwrite(data.table(col_processed = col), progress_path, append = TRUE)
  return(list(colname = col_title, type = detected_type))
}

results <- foreach(col = start_col:total_cols, .combine = 'rbind', .packages = 'data.table') %dopar% {
  process_column(col)
}

results <- as.data.table(results)
fwrite(results, "column_types_summary.csv")

print(results)
```

This code block sets up for data visualization by reading a summary file and preparing a pie chart to visualize the distribution of data types.

```{r data statistics(plot), message=FALSE}
file_path <- "column_types_summary.csv" 
data <- fread(file_path)


type_counts <- table(data$type)
type_counts_df <- as.data.frame(type_counts)
names(type_counts_df) <- c("Type", "Frequency")
type_counts_df$Prop <- type_counts_df$Frequency / sum(type_counts_df$Frequency)


if (nrow(type_counts_df) > 1) {
  pie_colors <- rainbow(nrow(type_counts_df))
  png(file = "type_distribution_pie_chart.png")
  pie(type_counts_df$Frequency, labels = paste0(type_counts_df$Type, ": ", round(type_counts_df$Prop * 100, 1), "%"), 
      col = pie_colors, main = "Pie Chart of Data Types")
  dev.off()
} else {
  cat("Not enough data to plot a pie chart.\n")
}
```

# Data Processing

Read subset for toy analysis

```{r subset ready and preprocess, message=FALSE}
data <- fread('ukbb_data/1000_rows.csv', na.strings = c("", "NA", "na"))
data <- convert_dates_to_numeric(data)
data <- apply_label_encoding(data)
data <- normalize_df(data)
data <- data %>% select(which(colSums(is.na(data)) != nrow(data)) %>% names())  
data <- missForest(data, verbose = TRUE)$ximp
data <- unique(data)                                  # remove duplication
data_matrix <- as.matrix(data)            # make sure the data is numeric
data <- data[, .SD, .SDcols = sapply(data, function(x) !all(x == 0))]
```

# Analysis

## K-means

This section performs a K-means clustering analysis. It identifies the optimal number of clusters using silhouette scores and show visualization of the of the K index parameter versus

```{r kmeans, warning=FALSE}
library(cluster)    
library(factoextra)
library(ggplot2)  # for plotting

set.seed(123) 

optimal_k <- 2
highest_silhouette <- 0
sil_scores <- numeric(29)  # Vector to store average silhouette scores for each k

for(k in 2:30) {
  km.res <- kmeans(data_matrix, centers = k, nstart = 25)
  ss <- silhouette(km.res$cluster, dist(data_matrix))
  avg.sil <- mean(ss[, 3])
  sil_scores[k-1] <- avg.sil  # Store the silhouette score
  
  if(avg.sil > highest_silhouette){
    highest_silhouette <- avg.sil
    optimal_k <- k
  }
}

clusters <- kmeans(data_matrix, centers = optimal_k, nstart = 25)

cat("Optimal number of clusters:", optimal_k, "\n")
print(clusters)

# Plotting the silhouette score vs. number of clusters
k_values <- 2:30
sil_plot <- ggplot(data = data.frame(k_values, sil_scores), aes(x = k_values, y = sil_scores)) +
            geom_line() +
            scale_x_continuous(breaks = 2:30) +
            theme_minimal() +
            ggtitle("Silhouette Score vs. Number of Clusters") +
            xlab("Number of Clusters (k)") +
            ylab("Average Silhouette Score")

print(sil_plot)
```

## Dimensional reduction with PCA，t-SNE and UMAP

Here, PCA, t-SNE, and UMAP techniques are applied for dimensionality reduction. The results are prepared for subsequent visualization.

```{r dimensionality-reduction}
# PCA
non_constant_data_matrix <- data_matrix[, apply(data_matrix, 2, var) != 0]
pca_result_2d <- prcomp(non_constant_data_matrix, scale. = TRUE, center = TRUE)
pca_2d <- pca_result_2d$x[, 1:2]  
pca_3d <- pca_result_2d$x[, 1:3]  

# t-sne
tsne_2d <- Rtsne(data_matrix, dims = 2, perplexity=20, verbose=TRUE)    
tsne_3d <- Rtsne(data_matrix, dims = 3, perplexity=20, verbose=TRUE)  

# umap
umap_result_2d <- umap(data_matrix, n_components = 2)                  
umap_result_3d <- umap(data_matrix, n_components = 3)                  
```

## PCA 2D and 3D Visualization

```{r pca visualization}
pca_2d_df <- as.data.frame(pca_2d)               # PCA 2d
pca_2d_df$cluster <- as.factor(clusters$cluster)
ggplot(pca_2d_df, aes(x = PC1, y = PC2, color = cluster)) + geom_point() + theme_minimal() + ggtitle("PCA 2D Visualization with Cluster Coloring")
pca_3d_df <- as.data.frame(pca_3d)
pca_3d_df$cluster <- as.factor(clusters$cluster)
fig_pca_3d_cluster <- plot_ly(pca_3d_df, x = ~PC1, y = ~PC2, z = ~PC3, color = ~cluster, colors = RColorBrewer::brewer.pal(3,"Dark2"), type = 'scatter3d', mode = 'markers')
fig_pca_3d_cluster
```

## t-SNE 2D and 3D Visualization

```{r tsne-visualization}
data_2d <- as.data.frame(tsne_2d$Y)               # t-sne 2d 
data_2d$cluster <- as.factor(clusters$cluster)    # add cluster result
data_3d <- as.data.frame(tsne_3d$Y)
data_3d$cluster <- as.factor(clusters$cluster)
ggplot(data_2d, aes(x = V1, y = V2, color = cluster)) + geom_point() + theme_minimal() +   ggtitle("t-SNE 2D Visualization with Cluster Coloring")
fig <- plot_ly(data_3d, x = ~V1, y = ~V2, z = ~V3, color = ~cluster, colors = RColorBrewer::brewer.pal(3,"Dark2"), type = 'scatter3d', mode = 'markers')
fig
```

## UMAP 2D and 3D Visualization

```{r umap-visualization}
data_3d <- as.data.frame(tsne_3d$Y)
data_3d$cluster <- as.factor(clusters$cluster)
data_2d <- data.frame(X = umap_result_2d$layout[,1], Y = umap_result_2d$layout[,2])  # umap 2d
data_2d$cluster <- as.factor(clusters$cluster)
ggplot(data_2d, aes(x = X, y = Y, color = cluster)) + 
  geom_point() + 
  theme_minimal() + 
  ggtitle("UMAP 2D Visualization with Cluster Coloring")
data_3d <- data.frame(X = umap_result_3d$layout[,1], Y = umap_result_3d$layout[,2], Z = umap_result_3d$layout[,3])
data_3d <- data.frame(X = umap_result_3d$layout[,1], Y = umap_result_3d$layout[,2], Z = umap_result_3d$layout[,3])  
data_3d$cluster <- as.factor(clusters$cluster)

fig <- plot_ly(data_3d, x = ~X, y = ~Y, z = ~Z, color = ~cluster, colors = RColorBrewer::brewer.pal(length(unique(data_3d$cluster)),"Dark2"), type = 'scatter3d', mode = 'markers')
fig

```
# Feature selection


## t-tests
This code performs feature selection by conducting pairwise t-tests between clusters for each feature in the data matrix, adjusting the p-values using the Benjamini-Hochberg method to identify statistically significant features. It creates a summary table to display significant features and their adjusted p-values.

```{r t test}
results_ttest <- list()  
feature_names <- colnames(data_matrix)  

unique_clusters <- unique(clusters$cluster)

for(i in 1:ncol(data_matrix)) {
  feature_results <- list()  
  
  for(j in 1:(length(unique_clusters)-1)) {
    for(k in (j+1):length(unique_clusters)) {
      
      group1 <- data_matrix[clusters$cluster == unique_clusters[j], i]
      group2 <- data_matrix[clusters$cluster == unique_clusters[k], i]
      
      if(var(group1) != 0 && var(group2) != 0) {
        feature_results[[paste(unique_clusters[j], unique_clusters[k], sep="_")]] <- t.test(group1, group2)$p.value
      } else {
        feature_results[[paste(unique_clusters[j], unique_clusters[k], sep="_")]] <- NA
      }
    }
  }
  
  results_ttest[[feature_names[i]]] <- feature_results
}

all_p_values <- sapply(results_ttest, function(feature) {
  min_p_value <- min(unlist(feature), na.rm = TRUE)
  return(ifelse(is.infinite(min_p_value), NA, min_p_value))
})

p_adjusted <- p.adjust(all_p_values, method = "BH")
significant_features <- which(p_adjusted < 0.05)

summary_table <- data.frame(
  Feature = feature_names[significant_features],
  P_Value = p_adjusted[significant_features]
)

print(summary_table)

```

## Boruta based random forest
This script applies the Boruta algorithm to identify important features in the data matrix, visualizing their importance using box-and-whisker plots, and then confirms the selection of significant features using the TentativeRoughFix function.

```{r random forest}
library(Boruta)
library(plotly)
library(tidyr)
library(dplyr)

cluster_numbers <- clusters$cluster
cluster_factors <- as.factor(cluster_numbers)

set.seed(123) # 
data_matrix_df <- as.data.frame(data_matrix)

boruta_output <- Boruta(cluster_factors ~ ., data=data_matrix_df, doTrace = 0)

library(plotly)
# plot(als, xlab="", xaxt="n")
# lz<-lapply(1:ncol(als$ImpHistory), function(i)
# als$ImpHistory[is.finite(als$ImpHistory[, i]), i])
# names(lz)<-colnames(als$ImpHistory)
# lb<-sort(sapply(lz, median))
# axis(side=1, las=2, labels=names(lb), at=1:ncol(als$ImpHistory), cex.axis=0.5, font = 4)

df_long <- tidyr::gather(as.data.frame(boruta_output$ImpHistory), feature, measurement)

plot_ly(df_long, x=~feature, y = ~measurement, color = ~feature, type = "box") %>%
    layout(title="Box-and-whisker Plots across all features (UKBB Data)",
           xaxis = list(title="Features", categoryorder = "total descending"), 
           yaxis = list(title="Importance"), showlegend=F)

select_result <- TentativeRoughFix(boruta_output)
print(select_result)
select_result$finalDecision
```

```{r knockoff}
library(knockoff)

results <- knockoff.filter(data, as.numeric(cluster_factors), fdr = 0.80)
print(results)

```

